---
title: "Blood pressure and Eosinophil count in African Ancestry"
author: "Uganda MR Hackathon Group3"
date: "2023-06-30"
output: html_document
---

#Load needed packaages
```{r}
library(plyr) 
library(TwoSampleMR)
library(devtools)
library(calibrate)
library(ggrepel)
library(ggthemes)
library(TwoSampleMR)
library(MRInstruments)
library(ggplot2)
library(png)
```

#Read in the Diastolic blood pressure snp file
```{r}
dbp_exp_dat <- read_exposure_data(
    filename = "diastolic_5e-5.txt",
    sep = "\t",
    snp_col = "variant_id",
    beta_col = "beta",
    se_col = "standard_error",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
    eaf_col = "effect_allele_frequency",
    pval_col = "p_value",
    samplesize_col = "n",
    phenotype_col = "Phenotype"
)
```


#Read in the systolic blood pressure snp file
```{r}
sbp_exp_dat <- read_exposure_data(
    filename = "systolic_5e-5.txt",
    sep = "\t",
    snp_col = "variant_id",
    beta_col = "beta",
    se_col = "standard_error",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
    eaf_col = "effect_allele_frequency",
    pval_col = "p_value",
    samplesize_col = "n",
    phenotype_col = "Phenotype"
)
```

#merge both into one dataframe
#Restrict using pvalue 5e-7 and clump snps
```{r}
merged_exp_dat = rbind(dbp_exp_dat, sbp_exp_dat)
merged_exp_dat = subset(merged_exp_dat, pval.exposure < 5e-7)
merged_exp_dat = clump_data(merged_exp_dat, pop="AFR")
```

#plot allele frequencies in Uganda population and Durban population from Outcome data
```{r}
Eosino_data = read.table(file = "eosinophil.txt", header = TRUE, sep = "\t")
x_column_name <- "eaf_uganda"
y_column_name <- "eaf_DDS"

# Fit linear regression model
lm_model <- lm(formula = paste(y_column_name, "~", x_column_name), data = Eosino_data)

# Extract R-squared value
r_squared <- summary(lm_model)$r.squared

# Create the scatter plot
scatter_plot <- ggplot(Eosino_data, aes(x = !!as.name(x_column_name), y = !!as.name(y_column_name))) +
  
  geom_point() +
  
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +  # Dotted line for x=y
  
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Regression line
  
  labs(title = "Scatter Plot",
       
       x = x_column_name,
       
       y = y_column_name,
       
       caption = paste0("R-squared = ", format(r_squared, digits = 4)))  # Display R-squared on the graph



# Display the scatter plot
png("UGA-DDS_Scatterplot.png")
   print(scatter_plot)
dev.off()
```

#extract exposure snps from outcome data
```{r}
outcome_dat_uga <- read_outcome_data(
    snps = merged_exp_dat$SNP,
    filename = "eosinophil.txt",
    sep = "\t",
    snp_col = "variant_id",
    beta_col = "beta_fe",
    se_col = "se_fe",
    effect_allele_col = "ea",
    other_allele_col = "oa",
    eaf_col = "eaf_uganda",
    pval_col = "pval_fe",
    samplesize_col = "n",
    phenotype_col = "Phenotype"
)
```
# Harmonise the data
```{r}
dat <- harmonise_data(merged_exp_dat, outcome_dat_uga, action = 2)
```
# Check the casual effect of DBP/SBP on Eosinophil count
```{r}
dat_dbp = subset(dat, exposure == "Diastolic_bp")
dat_sbp = subset(dat, exposure == "Systolic")
mr_results_sbp <- mr(dat_sbp, method_list=c("mr_ivw","mr_egger_regression","mr_weighted_median", "mr_weighted_mode"))
mr_results_sbp
results_sbp<-cbind.data.frame(mr_results_sbp$outcome,mr_results_sbp$nsnp,mr_results_sbp$method,mr_results_sbp$b,mr_results_sbp$se,mr_results_sbp$pval)
mr_results_dbp <- mr(dat_dbp, method_list=c("mr_ivw","mr_egger_regression","mr_weighted_median", "mr_weighted_mode"))
mr_results_dbp
results_dbp<-cbind.data.frame(mr_results_dbp$outcome,mr_results_dbp$nsnp,mr_results_dbp$method,mr_results_dbp$b,mr_results_dbp$se,mr_results_dbp$pval)

```
#Export the results
```{r}
write.csv(results,"DBP-SBP_Eosi_results.csv")
```

#Run sensitivity analyses

#Is there evidence of heterogeneity in the genetic effects?
```{r}
het <- mr_heterogeneity(dat)
het
```
#is there evidence of pleiotropy
```{r}
pleio <- mr_pleiotropy_test(dat)
pleio
res_single_sbp <- mr_singlesnp(dat_sbp)
res_single_dbp <- mr_singlesnp(dat_dbp)
```
#visualise the causal effect
```{r}
png("SBP_Eosi_scatter.png")
  mr_scatter_plot(mr_results_sbp, dat_sbp)
dev.off()
png("DBP_Eosi_scatter.png")
  mr_scatter_plot(mr_results_dbp, dat_dbp)
dev.off()
```
#Generate a forest plot of each of the SNP effects, which are then meta-analysed using the #IVW and MR-Egger methods
```{r}
png("SBP_Eosi_forest.png")
  mr_forest_plot(res_single_sbp)
dev.off()
png("DBP_Eosi_forest.png")
  mr_forest_plot(res_single_dbp)
dev.off()
```
#Generate a funnel plot to check for asymmetry

```{r}
png("SBP_Eosi_funnel.png")
  mr_funnel_plot(res_single_sbp)
dev.off()
png("DBP_Eosi_funnel.png")
  mr_funnel_plot(res_single_dbp)
dev.off()
```
#Run a leave-one-out analysis and generate a plot to test whether any one SNP is driving any #pleiotropy or asymmetry in the estimates

```{r}
res_loo_sbp <- mr_leaveoneout(dat_sbp)
png("SBP_Eosi_loo.png")
  mr_leaveoneout_plot(res_loo_sbp)
dev.off()
res_loo_dbp <- mr_leaveoneout(dat_dbp)
png("DBP_Eosi_loo.png")
  mr_leaveoneout_plot(res_loo_dbp)
dev.off()
```